Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-21 probabilisticProgrammingMartingales.tex
Question about the example that illustrates that the antitone function must be everywhere defined (Example 4).

\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$).
However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$. So I propose the following.}

@@

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
\step[=]{definition of expectation conditioning on an \emph{event}}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\[
\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
\]
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
\end{proof}

Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values). Consider, for example, a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T = $ the least $n$ such that $2^{-n} < X$, $Y_k = 4-2^{-k}$ if $k < T$, or $0$ otherwise, and $\epsilon(x) = \frac{4-x}{4}$. In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. It is an antitone strict supermartingale except that $\epsilon$ goes to $0$ at $4$, which is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $T = \infty$ with probability $\frac 1 2$.

@@

  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions. Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$. For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions. In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction). For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial. If $R$ is a $tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded. Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
\end{proof}

@@

If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved. Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi n$:
\begin{align*}
\Theta[n] \mapsto n \\
\Xi 10 \mapsto 12.
\end{align*}
For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$. These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
\end{example}

>>>>

%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$). However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}

@@

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
\step[=]{definition of expectation conditioning on an \emph{event}}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\[
\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
\]
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
\end{proof}

Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).

\begin{example}
Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
\[
Y_k := 
\begin{cases}
4-2^{-k} & \hbox{if $k < T$}\\
0 & \hbox{otherwise}
\end{cases} 
\]
and $\epsilon(x) = \frac{4-x}{4}$. 
In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. It is an antitone strict supermartingale \changed[akr]{except that $\epsilon$ goes to $0$ at $4$, which is larger than any value $(Y_n)_{n \in \omega}$ can actually take}, and $T = \infty$ with probability $\frac 1 2$.
\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?}
\end{example}

@@

  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions. Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$. For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions. In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction). For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial. If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded. Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
\end{proof}

@@

If instead, we take the ranking function to be defined after the $\tY$-reduction but before the $\tsample$-reduction as well, this can be resolved. Letting $\Theta[n] = \underline 0 \oplus_{\frac 1 n} \Xi \, n$:
\begin{align*}
\Theta[n] &\mapsto n \\
\Xi \, 10 & \mapsto 12.
\end{align*}
For each $n$, $\Theta[n]$ reduces to either $0$ or $\Theta[n+1]$, with a $\tY$-reduction only in the latter case, and the condition that this is an antitone partial ranking function is that $n \geq \frac{(n-1)(n+1)}{n} + \frac{n-1}{n} \epsilon(n+1)$ and $12 \geq 11 + \epsilon(11)$. These are satisfied by setting $\epsilon(x) = \min(1, 1/x)$, therefore this term is AST.
\end{example}
